---
title: "PLSC 30600 - Week 7 Discussion Section"
output:
  xaringan::moon_reader:
    self_contained: true
    mathjax: "default"
    css: [beamerish.css]
    nature:
      highlightLines: true
      ratio: '16:9'
      countIncrementalSlides: false
---

# Today

- Situating the IV estimator

- Inference: randomization versus asymptotic

- Caching


---

# Before there was IV

$$Y_i=\alpha + \tau D_i + \epsilon_i$$

--

We can use OLS to estimate $\beta$:

$$\hat{\tau}_{OLS} \rightarrow \tau$$

--

... but only under several conditions

- OLS assumptions: random sampling, finite moments, full rank
- Structural/causal interpretation assumption: $E[D_ie_i]=0$

--

When D is endogenous, or not randomly assigned, $D$ and $e_i$ correlate, meaning $E[D_ie_i] \neq 0$.

Can't estimate $\tau$ with OLS.

--

Solution: 

$$\hat{\tau}_{IV} \rightarrow \tau$$

---

# Enters the IV estimator

$$\hat{\tau}_{IV} = \frac{Cov(Y,Z)}{Cov(D,Z)}$$

--

And when the treatment is binary, the Wald estimator:

$$\hat{\tau}_{Wald} = \frac{E[Y|Z=1]-E[Y|Z=0]}{E[Y|D=1]-E[Y|D=0]}$$

--

How to get $\hat{\tau}_{IV}$: 2SLS (OLS twice)

--

With identifying assumptions, we can use IV estimators (and therefore 2SLS) to recover LATE.

---

# Spot the identifying assumptions

How many assumptions for $\hat{\tau}_{IV} \rightarrow \tau$ relate to the assumptions or forms we've already specified?

$$Y_i=\alpha + \tau D_i + \epsilon_i$$

$$\hat{\tau}_{IV} = \frac{Cov(Y,Z)}{Cov(D,Z)}$$

$$E[Z_ie_i]=0$$

---

# What about inference?

We have so far talked about estimation, that we can use IV/2SLS to recover LATE under certain assumptions.

--

For inference, such as testing $H_0: \tau=\tau_0$, we need to know what the distribution of $\hat\tau-\tau_0$ is

--

This means we need to know the distribution of $\hat{\tau}$.

How do we do this?

---

# Asymptotic versus randomized inference

There are two ways to get the distribution of $\hat{\tau}$:

--

**Asymptotically**

Intuition:
- The problem: We have one sample from the population, and therefore one $\hat{\tau}$.
- The solution: We redraw samples.

--

$$\sqrt n(\hat\tau-\tau_0) \rightarrow_d N(0,\sigma^2)$$

We don't know $\sigma^2$ either, so we estimate it using the data we have.

--

For IV/2SLS:

$$Var(\hat{\tau}_{2SLS})=\frac{\hat{\sigma^2}}{n}$$

---

# Asymptotic versus randomized inference

There are two ways to get the distribution of $\hat{\tau}$:

--

**Randomization inference**

Intuition:
- The problem: We observe one treatment assignment, and therefore one $\hat{\tau}$.
- The solution: We reshuffle treatment.

--

Permutation test for IV:
- Enumerate all possible assignments of Z
- Compute $\hat\tau$ under each assignment
- Obtain the entire distribution of $\hat\tau$

---

# Asymptotic versus randomized inference

Difference: the source of uncertainty

i.e., What is treated as random?

--

> Asymptotic inference: sampling

> Randomized inference: design

---

# Two types of inferences, two traditions

**Fisher**
- $H_0$: often sharp null ($\tau=0$)
- p-value: How surprising is $\hat\tau$ under $H_0$? 
- Report p-value as evidence against $H_0$

--

**Neyman-Pearson**
- Specify $H_0$ and $H_1$
- Specify the decision rule: type 1 ($\alpha$) and type 2 ($\beta$) errors
- Reject $H_0$ when $\tau$ falls in a specific rejection region

--

**Different emphases**
- Fisher: Exact distribution of test statistic
- Neyman-Pearson: Long-run behavior, average effects

---

# “The null ritual”: Gigerenzer, Krauss & Vitouch (2004)

Traditional hypothesis testing workflow conflates the two views:

1. Set up a null $H_0$: “no difference/no effect” (Fisher)
2. Choose $\alpha = 0.05$ (Neyman-Pearson)
3. Compute a p-value (Fisher)
4. If $p < 0.05$, declare “significant” and treat $H_0$ as false (Neyman-Pearson)
5. If $p \ge 0.05$, declare “no effect” (Fisher)

--

What this does
- Conflates evidence (Fisher) with decisions (N–P)
- Encourages dichotomous thinking
- Often ignores effect sizes, uncertainty, and design

---

```{r setup-sim, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
```

# Design-based variation

Simulate a set of potential outcomes and recover the true ATE

```{r finite-pop}
N <- 2000
X <- rnorm(N)

tau <- 2
Y0 <- 1 + 0.5*X + rnorm(N, sd = 1)
Y1 <- Y0 + tau
ATE_finite_true <- mean(Y1 - Y0)
ATE_finite_true
```

---

# Design-based variation

Randomize treatment (complete randomization, half treated), estimate diff-in-means.

```{r one-experiment}
W <- rep(0, N)
W[sample.int(N, N/2)] <- 1
Yobs <- W*Y1 + (1-W)*Y0
mean(Yobs[W==1]) - mean(Yobs[W==0])
```

---

# Design-based sampling distribution

Obtain the distribution of the differences-in-means estimator by permuting through different assignment mechanisms.

```{r design-dist, cache=TRUE}
R <- 2000
ate_hat_design <- replicate(R, {
  W <- rep(0, N)
  W[sample.int(N, N/2)] <- 1
  Yobs <- W*Y1 + (1-W)*Y0
  mean(Yobs[W==1]) - mean(Yobs[W==0])
})

c(mean = mean(ate_hat_design), sd = sd(ate_hat_design))
```

---

# Design-based sampling distribution

Visualize

```{r design-plot, cache=TRUE}
hist(ate_hat_design,
     main = "Design-based variation (population fixed)",
     xlab = "ATE estimator (diff-in-means)")
abline(v = ATE_finite_true, lwd = 3)
```

---

# Sampling-based variation

Same distribution, but this time simulate a super-population

```{r}
N_pop <- 5000
X_pop  <- rnorm(N_pop)
tau <- 2
Y0_pop <- 1 + 0.5 * X_pop + rnorm(N_pop, sd = 1)
Y1_pop <- Y0_pop + tau
```

---

# Sampling-based sampling distribution

Draw iid samples from the super-population and compute the ATE for each sample

```{r}
ate_hat_super <- replicate(R, {
  idx_sample <- sample.int(N_pop, N, replace = FALSE)
  Y0 <- Y0_pop[idx_sample]
  Y1 <- Y1_pop[idx_sample]
  W <- rbinom(N, 1, 0.5)
  Yobs <- W * Y1 + (1 - W) * Y0
  mean(Yobs[W == 1]) - mean(Yobs[W == 0])
})

c(mean = mean(ate_hat_super), sd = sd(ate_hat_super))
```

---

# Sampling-based variation

Visualize

```{r super-plot, cache=TRUE}
hist(ate_hat_super,
     main = "Sampling-based variation\n(new sample each time)",
     xlab = "Diff-in-means ATE estimator")
abline(v = tau, lwd = 3)
```

---

# Bootstrap is slow (computational repetition)

Create **one realized dataset** and bootstrap the ATE.

```{r realized-dataset}
set.seed(999)
N3 <- 3000
X3 <- rnorm(N3)
tau3 <- 2
Y0_3 <- 1 + 0.5*X3 + rnorm(N3, sd = 1)
Y1_3 <- Y0_3 + tau3

W3 <- rbinom(N3, 1, 0.5)
Y3 <- W3*Y1_3 + (1-W3)*Y0_3

ate_obs <- mean(Y3[W3==1]) - mean(Y3[W3==0])
ate_obs
```

---

# Cache method 1: knitr chunk caching

`{r cache=TRUE}`

```{r boot-knitr-cache, cache=TRUE}
B <- 4000
ate_boot <- replicate(B, {
  idx <- sample.int(N3, N3, replace = TRUE)
  mean(Y3[idx][W3[idx]==1]) - mean(Y3[idx][W3[idx]==0])
})

c(se = sd(ate_boot),
  ci_lo = unname(quantile(ate_boot, 0.025)),
  ci_hi = unname(quantile(ate_boot, 0.975)))
```

---

# Cache method 1: knitr chunk caching

Great when:
- you’re re-knitting the **same** document while editing
- the slow chunk’s inputs are stable

But, beware:
- it’s tied to a cache folder (e.g., `*_cache/` or `knitr_cache/`)
- changes in previous chunks will not apply

---

# Cache method 2: caching to `.rds`

```{r boot-rds-cache}
rds_file <- "ate_bootstrap_draws.rds"

if (file.exists(rds_file)) {
  ate_boot_rds <- readRDS(rds_file)
} else {
  B <- 4000
  ate_boot_rds <- replicate(B, {
    idx <- sample.int(N3, N3, replace = TRUE)
    mean(Y3[idx][W3[idx]==1]) - mean(Y3[idx][W3[idx]==0])
  })
  saveRDS(ate_boot_rds, rds_file)
}

c(se = sd(ate_boot_rds),
  ci_lo = unname(quantile(ate_boot_rds, 0.025)),
  ci_hi = unname(quantile(ate_boot_rds, 0.975)))
```

---

# Cache method 2: caching to `.rds`

Especially great when:
- results are expensive enough to be a *data artifact* (big sims/bootstraps)
- you want reuse across documents/sessions
- you want robustness to cache deletion

--

Thesis:

> `cache=TRUE` speeds up knitting.  
> `.rds` makes computation portable.

---

# References

- Aronow, P. M., & Miller, B. T. (2019) *Foundations of Agnostic Statistics*

- Gigerenzer, G., Krauss, S., & Vitouch, O. (2004). “The null ritual.” In *The Sage handbook of quantitative methodology for the social sciences*, 391–408.

- Xie, Y. Cache time-consuming code chunks. R Markdown Cookbook. https://yihui.org/rmarkdown-cookbook/cache
