---
title: "Section W8 - DiD to TWFE"
author: "Your Name Here"
date: "2026-02-26"
output: html_document
---

```{r}
library(tidyverse)
library(haven)
library(estimatr)
library(fixest)
library(panelView)
```

# 0. Set-up

```{r}
ck_url <- "https://raw.githubusercontent.com/UChicago-pol-methods/PLSC30600/main/data/ck_minwage_did.csv"
paglayan_url <- "https://raw.githubusercontent.com/UChicago-pol-methods/PLSC30600/main/data/paglayan-2019.dta"

minwage <- read_csv(ck_url)

union <- haven::read_dta(paglayan_url) %>%
  filter(!is.na(studteachratio),
         State != "DC", State != "WI",
         year > 1959, year < 1997)
```

# 1. Warm-Up: Card & Krueger (1994)

Recall the context: New Jersey (`STATE==1`) raised its minimum wage in April 1992; Pennsylvania (`STATE==0`) did not. Card and Krueger surveyed fast-food restaurants *before* (Wave 1) and *after* (Wave 2) to see how the minimum wage raise affects full time employment (`EMPFT` for wave 1; `EMPFT2` for wave 2).

## 1.1 What is in the naive DiD estimator?

Recall:

$$\tau_{DiD}\equiv\big(E[Y \mid G = 1, t = 1]-E[Y \mid G = 1, t = 0]\big)-\big(E[Y \mid G = 0, t = 1]-E[Y \mid G = 0, t = 0]\big)$$
**What does each of the four expected value mean? What about the first difference and the second difference?**

**What is the parallel trends assumption here?**

# 1.2 Estimate TWFE  (`feols()`)

Hint: you may need to restructure the data

```{r}

```

# 1.3 Confirm TWFE matches DiD estimate

Run the following code that calculates each component of the DiD equation separately to obtain the estimate. Make sure your TWFE results match.

```{r}
minwage %>%
  pivot_longer(cols = c(EMPFT, EMPFT2),
               names_to = "wave", values_to = "Y") %>%
  mutate(time  = if_else(wave == "EMPFT", 0, 1),
         treat = if_else(STATE == 1, "NJ (treated)", "PA (control)")) %>%
  mutate(
    G  = if_else(treat == "NJ (treated)", 1, 0),
    D  = G * time,
    id = rep(1:(nrow(minwage)), each = 2)
  ) %>% 
  group_by(treat, time) %>%
  summarise(mean_Y = mean(Y, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(names_from = time, values_from = mean_Y,
              names_prefix = "t") %>%
  mutate(change = t1 - t0) %>% 
  summarise(did =
              change[treat == "NJ (treated)"] -
              change[treat == "PA (control)"]) %>%
  pull(did)
```

# 2. Staggered Adoption with Paglayan (2019)

Paglayan (2019) studies whether laws requiring collective bargaining (`CBrequired_SY`) with teachers increase public education spending, as measured with logged per-pupil expenditures `lnppexpend`.

Using staggered adoption of these laws across U.S. states, she applies a difference-in-differences design with state and year fixed effects to compare outcomes before and after adoption.

**What is the parallel trends assumption here?**

## 2.1 Visualize the staggered treatment timing using panelview()

**What are the treatment cohorts (groups)?**

```{r}

```

## 2.2 Estimate the naive TWFE using feols()

```{r}


```

## 2.3 What are the hidden 2x2s?

Below, I grouped treatment into two cohorts: early and late. Referencing slide 48 in Week 8's slides, we know that TWFE effectively combines multiple 2x2 DiD comparisons. In the case of two treatment cohorts, there are four:

+ Early vs never
+ Late vs never
+ Early vs late
+ Late vs early

**In plain words, why can't we compare late vs. early?**

Using the example of `est1` and the helper function, separate these four 2x2s. Compare them to the naive TWFE estimate.

```{r}
# Define early/late/never (coarse timing groups)
median_yr <- median(union$YearCBrequired, na.rm = TRUE)

union_gb <- union |>
  mutate(group_label = case_when(
    is.na(YearCBrequired)       ~ "never",
    YearCBrequired <= median_yr ~ "early",
    TRUE                        ~ "late"
  ))

all_yrs <- sort(unique(union_gb$year))

# Pick a single "early adoption time" and "late adoption time" to anchor the periods
g_early <- min(union_gb$YearCBrequired[union_gb$group_label == "early"], na.rm = TRUE)
g_late  <- min(union_gb$YearCBrequired[union_gb$group_label == "late"],  na.rm = TRUE)
```

```{r}
# Helper function to do individual 2x2s
did_2x2 <- function(df, g_treat, g_ctrl, pre_t, post_t) {
  yt_pre  <- mean(df$lnppexpend[df$group_label == g_treat & df$year %in% pre_t],  na.rm=TRUE)
  yt_post <- mean(df$lnppexpend[df$group_label == g_treat & df$year %in% post_t], na.rm=TRUE)
  yc_pre  <- mean(df$lnppexpend[df$group_label == g_ctrl  & df$year %in% pre_t],  na.rm=TRUE)
  yc_post <- mean(df$lnppexpend[df$group_label == g_ctrl  & df$year %in% post_t], na.rm=TRUE)
  (yt_post - yt_pre) - (yc_post - yc_pre)
}
```

```{r}
# Early vs never
est1 <- did_2x2(union_gb, "early", "never",
                pre_t  = all_yrs[all_yrs < g_early],
                post_t = all_yrs[all_yrs >= g_early])

# Late vs never
est2 <- 

# Early vs late 
est3 <- 

# Late vs early
est4 <- 
  
```


## 2.4 Illustration (no coding): TWFE is a weighted average of 2x2 comparisons

```{r}
#install.packages("bacondecomp")
library(bacondecomp)

bacon_results <- bacon(
  lnppexpend ~ CBrequired_SY,
  data = union,
  id_var = "State",
  time_var = "year"
)

head(bacon_results)

weighted_avg <- sum(bacon_results$weight * bacon_results$estimate)
weighted_avg #same as naive TWFE
```


## 2.5 Robust estimation with staggered adoption using sunab()

Estimate a robust event study with k=-1 as the reference period and plot it. Hint: you may need to restructure the data.

```{r}

```

